Can We Crunch the Social Graph on the Cheap?

### 2011-01-25 doc/hotpar2011/softxmt-hotpar2011.tex
added small fixes and references up to 3.2.1

The most interesting computational challenge comes from large,
low-diameter, power law graphs: this combination makes extracting
performance difficult. Their size keeps them from fitting in a single
commodity machine's memory. Their low diameter means that they are
difficult to lay out with locality: every node needs to be near every
other node. Their power law distribution means that a few vertices
generate much more work than the rest, leading to computational
imbalance and even less locality. Together, these properties lead to a
difficult conclusion: each edge traversal is likely to be a very
expensive cache miss.

Multithreading is a technique that has been used successfully to
implement efficient computations for these graphs. The Cray XMT is an
example of such an approach: it solves the memory latency problem
through concurrency rather than caching. Each XMT processor supports
128 hardware contexts and 1024 outstanding memory operations, and is
able to switch contexts every cycle. This ability comes at a cost: the
XMT is an expensive, non-commodity machine.

@@

Whereas in the PGAS model coherency of global copies is implictly
maintained across the system, we forego caching of globally shared
data.  A node can exploit temporal locality of data stored in
the global space only by first copying it into the node's local space.
In this way, a node can access data at random across the system while
incurring no hardware coherence overhead: consistency is enforced by
our mechanisms for explicit global synchronization.

@@

Thus, the programming model provides the programmer with options: express
locality within the private space, and it will be exploited to reduce
latency and bandwidth; express concurrency, and it will be used toward
tolerating latency and reducing the cost of bandwidth.

@@

We minimize context size by treating context switches as function
calls, as in~\cite{charm}. This allows us to save and restore the
minimum number of registers allowed by the ABI, and depend on the
compiler to save and restore any other registers in use. We minimize
context switch time both by keeping contexts small and by doing all
switching and scheduling in user space, as in \todo{cite other green
  threads packages}.

@@

\subsubsection{Single-node implementation}
We use prefetch instructions for the issue operation, and regular
blocking reads and writes for the complete operation. Since we expect
these global references to have low locality, we use the
``non-temporal'' form of prefetch, which tries to bring the data near
the core without disturbing the state of the cache
hierarchy. \todo{cite intel docs}

@@

We believe we will need hundreds of outstanding memory references per processor to
cover the latency of remote references in a multi-node
system. Unfortunately, commodity processors have much smaller limits
on memory concurrency: section \todo{cite evaluation} finds a limit of
36 concurrent loads. We will have to manage memory concurrency on our
own to avoid this limit. We describe approaches in both software and
hardware.

@@

\subsubsection{Base memory system performance}
\todo{title might be improved}

>>>>

The most interesting computational challenge comes from large,
low-diameter, power law graphs: this combination makes extracting
performance difficult. Their size keeps them from fitting in a single
commodity machine's memory. Their low diameter means that they are
difficult to lay out with locality: every vertex needs to be near every
other vertex. Their power law distribution means that a few vertices
generate much more work than the rest, leading to computational
imbalance and even less locality. Together, these properties lead to a
difficult conclusion: each edge traversal is likely to be a very
expensive cache miss.

Multithreading is a technique that has been used successfully to
implement efficient computations for these graphs. The Cray XMT is an
example of such an approach: it solves the memory latency problem
through concurrency rather than caching. Each XMT processor supports
128 hardware contexts and 1024 outstanding memory operations, and is
able to switch contexts every cycle \todo{cite XMT description}. This ability comes at a cost: the
XMT is an expensive, non-commodity machine. \todo{don't we want to also say XMT has poor performance on x kind of apps?}

@@

Whereas in the PGAS model coherency of global copies is implictly
maintained across the system \todo{cite pgas?}, we forego caching of globally shared
data.  A node can exploit temporal locality of data stored in
the global space only by first copying it into the node's local space.
In this way, a node can access data at random across the system while
incurring no hardware coherence overhead: consistency is enforced by
our mechanisms for explicit global synchronization.

@@

Thus, the programming model provides the programmer with options: express
locality within the private space, and it will be exploited to reduce
latency and bandwidth \todo{exploited by what? by the exisiting locality hardware like cache hierarchy?}; express concurrency, and it will be used toward
tolerating latency and reducing the cost of bandwidth.

@@

We minimize context size by treating context switches as function
calls, as in~\cite{charm}. This allows us to save and restore the
minimum number of registers allowed by the ABI, and depend on the
compiler to save and restore any other registers in use. We minimize
context switch time both by keeping contexts small and by doing all
switching and scheduling in user space, as in \todo{cite other green
  threads packages, like qthreads}.

@@

\subsubsection{Single-node implementation}
We use prefetch instructions for the issue operation, and regular
blocking reads and writes for the complete operation. Since we expect
these global references to have low locality, we use the
``non-temporal'' form of prefetch, which tries to bring the data near
the core without disturbing the state of the cache
hierarchy \cite{intel:swdev}.

@@

We believe we will need hundreds of outstanding memory references per processor to
cover the latency of remote references in a multi-node
system\todo{ can we make a stronger statement? we know it will be 300$+$ according to our calculations}. Unfortunately, commodity processors have much smaller limits
on memory concurrency: section~\ref{subsubsec:evalsinglebase} finds a limit of
36 concurrent loads. We will have to manage memory concurrency on our
own to avoid this limit. We describe approaches in both software and
hardware.

@@

\subsubsection{Base memory system performance}
\label{subsubsec:evalsinglebase}
\todo{title might be improved}
